%%This is a very basic article template.
%%There is just one section and two subsections.
\documentclass[a4paper]{article}
\usepackage{a4wide}
\usepackage{amsmath}

\begin{document}

\title{Probability and Statistics for Data Analysis\\Assignment 1}
\author{Charalampos Kaidos}

\maketitle

\begin{enumerate}

  \item Assume that $A$ and $B$ are events of the sample space $S$ for which we
  know: $2P(A)-P(A')=3/5$, $P(B|A)=5/8$, $P(A|B)=4/9$. Calculate the following
  probabilities:
    \begin{enumerate}
      \item $P(A)$
        \begin{align}
          2P(A)-P(A')=3/5& \Leftrightarrow 2P(A)-(1-P(A))=3/5 \nonumber\\
          & \Leftrightarrow 2P(A)-1+P(A)=3/5 \nonumber\\
          & \Leftrightarrow 3P(A)=8/5 \nonumber\\
          & \Leftrightarrow P(A)=8/15
        \end{align}
      \item $P(A\cap B)$
        \begin{align}
          P(B|A)=\frac{P(A\cap B)}{P(A)}& \Leftrightarrow P(A\cap
          B)=P(B|A)P(A) \nonumber \\
          & \Leftrightarrow P(A\cap B)=5/8\times8/15=1/3
        \end{align}
      \item $P(B)$
        \begin{align}
          P(A|B)=\frac{P(A\cap B)}{P(B)}& \Leftrightarrow P(B)=\frac{P(A\cap
          B)}{P(A|B)}=\frac{1/3}{4/9}=9/12=3/4
        \end{align}
      \item $P(A\cup B)$
        \begin{align}
          P(A\cup B)=P(A)+P(B)-P(A\cap B)=8/15+3/4-1/3=171/180
        \end{align}
      \item Are the events $A$ and $B$ independent? \\
      Two events $A$ and $B$ are independent if $P(A\cap B)=P(A)P(B)$. For our
      problem:
      $$P(A)P(B)=8/15\times3/4=2/5\neq P(A\cap B)$$
      Thus the events $A$ and $B$ are not independent.
    \end{enumerate}
  \item Suppose that a sample space $S$ has $n$ elements. Prove that the number
  of subsets that can be formed from the elements of $S$ is $2^n$. \\
  Each element of the sample space can either belong to a certain subset or not.
  Thus there are two ways to create each subset either by including each element
  or not. Thus there are $2^n$ subsets.
  \item Two players, A and B, alternatively and independently flip a coin and
  the first player to obtain a head wins. Assume player A flips first.
    \begin{enumerate}
      \item If the coin is fair, what is the probability that player A wins?\\
      The coin is fair means that $P(\text{heads})=P(\text{tails})=1/2$. Player
      A wins if he flips heads. As each flip is an indipendent event, this
      happens with probability $P(\text{heads})=1/2$ on the first flip, with
      probability $P(\text{tails,tails,head})=1/8$ on the third flip and with
      probability $P(\text{tails,tails,tails,tails,head})=1/32$. In general the
      probability that player A wins in his $n^{th}$ attemp is:
      $$P(A\text{ wins in } n^{th} \text{
      attempt})=\frac{1}{2\times2^{2(n-1)}}$$
      Thus the probability that a wins is:
      $$P(A \text{
      wins})=\sum_{i=1}^{\infty}\frac{1}{2\times2^{2(i-1)}}=\sum_{i=0}^{\infty}\frac{1}{2\times2^{2i}}$$
      \item More generally assume that $P(\text{head})=p$ (not necessarily
      $1/2$). What is the probability that player A wins?\\
      If $P(\text{heads})=p$ then $P(\text{tails})=1-p$. Based on the
      calculations above:
      $$P(A \text{ wins})=\sum_{i=0}^{\infty}\frac{1}{p\times(1-p)^{2i}}$$
      \item Show that $\forall p$ such that $0<p<1$, we have that $P(A\text{
      wins})>1/2$.
      $$P(B \text{ wins})=1-P(A \text{ wins})$$
      Also as player flips the coin second, he wins on his first attempt if
      player a flips a tails and then player B flips a heads. In general his
      probability to win is:
      $$P(B \text{ wins})=\sum_{i=0}^{\infty}\frac{1}{p\times(1-p)^{2i+1}}$$
      Obviously $P(A)>P(B)$ thus:
      $$P(A)>P(B) \Leftrightarrow P(A)>1-P(A) \Leftrightarrow P(A)>1/2$$
    \end{enumerate}
  \item A telegraph signals ``dot'' and ``dash'' sent in the proportion $3 : 4$,
  where erratic transmission cause a dot to become dash with probability $1/4$
  and a dash to become a dot with probability $1/3$.
    \begin{enumerate}
      \item If a dash is received, what is the probability that a dash has been
      sent?\\
      From the data we have we can get these information:
      $$P(\text{dot sent})=3/7$$
      $$P(\text{dash sent})=4/7$$
      $$P(\text{dash received}|\text{dot sent})=1/4$$
      $$P(\text{dot received}|\text{dot sent})=3/4$$
      $$P(\text{dot received}|\text{dash sent})=1/3$$
      $$P(\text{dash received}|\text{dash sent})=2/3$$
      The probability we are looking for is $P(\text{dash sent}|\text{dash
      received})$. From Bayes' rule:
      $$P(\text{dash sent}|\text{dash received})=P(\text{dash
      received}|\text{dash sent})\times\frac{P(\text{dash
      sent})}{P(\text{dash received})}$$
      Also:
      \begin{align}
      P(\text{dash received})&=P(\text{dash received}\cap\text{dash
      sent})+P(\text{dash received}\cap\text{dot sent})\nonumber\\
      &=P(\text{dash received}|\text{dash sent})P(\text{dash sent})+\nonumber\\
      &\quad+P(\text{dash received}|\text{dot sent})P(\text{dot sent})\nonumber\\
      &=2/3\times4/7+1/4\times3/7\nonumber\\
      &=8/21+3/28\nonumber\\
      &=41/84
      \end{align}
      Thus:
      $$P(\text{dash sent}|\text{dash
      received})=2/3\times\frac{4/7}{41/84}=\frac{672}{861}=\frac{32}{41}$$
      \item Assuming independence between signals, if the message dot-dot was
      received, what is the probability distribution of the four possible
      messages that could have been sent?\\
      Using the same reasoning as above we have:
      \begin{align}
      P(\text{dot received})&=P(\text{dot received}\cap\text{dot
      sent})+P(\text{dot received}\cap\text{dash sent})\nonumber\\
      &=P(\text{dot received}|\text{dot sent})P(\text{dot sent})+\nonumber\\
      &\quad+P(\text{dot received}|\text{dash sent})P(\text{dash
      sent})\nonumber\\
      &=3/4\times3/7+1/3\times4/7\nonumber\\
      &=9/28+4/21\nonumber\\
      &=43/84
      \end{align}
      There are two ways to get the 'dot'; $\text{dot received}\cap\text{dot
      sent}$ and $\text{dot received}\cap\text{dash sent}$. So to get the
      'dot-dot' message there are 4 ways. Since we know that dot was received
      we are interested in the conditional probabilities bellow.
      \begin{align}
      P(\text{dot sent}|\text{dot received})&=P(\text{dot received}|\text{dot
      sent})\times\frac{P(\text{dot sent})}{P(\text{dot received})}\nonumber\\
      &=3/4\times\frac{3/7}{43/84}\nonumber\\
      &=3/4\times252/301=756/1204=27/43
      \end{align}
      \begin{align}
      P(\text{dash sent}|\text{dot received})&=P(\text{dot
      received}|\text{dash sent})\times\frac{P(\text{dash sent})}{P(\text{dot
      received})}\nonumber\\
      &=1/3\times\frac{4/7}{43/84}=1/3\times336/301=336/903=16/43
      \end{align}
      Finally we are calculate the following 4 probabilities:
      \begin{align}
      P(\text{dot sent}_1|\text{dot received}_1\cap\text{dot sent}_2|\text{dot
      received}_2)&=P(\text{dot sent}_1|\text{dot received}_1)\times\nonumber\\
      &\quad\times P(\text{dot sent}_2|\text{dot received}_2)\nonumber\\
      &=27/43\times27/43=729/1849
      \end{align}
      \begin{align}
      P(\text{dash sent}_1|\text{dot received}_1\cap\text{dot sent}_2|\text{dot
      received}_2)&=P(\text{dash sent}_1|\text{dot received}_1)\times\nonumber\\
      &\quad\times P(\text{dot sent}_2|\text{dot received}_2)\nonumber\\
      &=16/43\times27/43=432/1849
      \end{align}
      \begin{align}
      P(\text{dot sent}_1|\text{dot received}_1\cap\text{dash sent}_2|\text{dot
      received}_2)&=P(\text{dot sent}_1|\text{dot received}_1)\times\nonumber\\
      &\quad\times P(\text{dash sent}_2|\text{dot received}_2)\nonumber\\
      &=27/43\times16/43=432/1849
      \end{align}
      \begin{align}
      P(\text{dash sent}_1|\text{dot received}_1\cap\text{dash sent}_2|\text{dot
      received}_2)&=P(\text{dash sent}_1|\text{dot received}_1)\times\nonumber\\
      &\quad\times P(\text{dash sent}_2|\text{dot received}_2)\nonumber\\
      &=16/43\times16/43=256/1849
      \end{align}
    \end{enumerate}
  \item Let $X$ be a continuous random variable with pdf $f(x)$ and cdf $F(x)$.
  For a fixed number $x_0$ (such that $F(x_0)<1$), define the function:
    \begin{align} g(x)=
      \begin{cases}
        \frac{f(x)}{1-F(x_0)} \quad &x\geq x_0 \\
        0 \quad &x<x_0
      \end{cases}
    \end{align}
  Prove that $g(x)$ is a pdf (also known as hazard function).\\
  For $g(x)$ to be a pdf the following must be true:
  $$g(x)\geq0\quad \text{for all }x$$
  and
  $$\int_{-\infty}^{\infty}g(x)dx=1$$
  Since $f(x)$ is a pdf then $f(x)\geq0$ for all $x$. Also since $F(x_0)<1$ then
  $1-F(x_0)>0$. Thus $\frac{f(x)}{1-F(x_0)}\geq0$. So $g(x)\geq0$ for all $x$.
  Moreover $\int_{-\infty}^{\infty}f(x)dx=1$ and
  $F(x)=\int_{-\infty}^{x}f(x)dx$. Thus $\int_{x_0}{\infty}f(x)dx=1-F(x_0)$.
  Finally:
  \begin{align}
  \int_{-\infty}^{\infty}g(x)dx&=\int_{x_0}^{\infty}g(x)dx=\int_{x_0}^{\infty}\frac{f(x)}{1-F(x_0)}dx\nonumber\\
  &=\frac{1}{1-F(x_0)}\times\int_{x_0}^{\infty}f(x)dx=\frac{1-F(x_0)}{1-F(x_0)}=1
  \end{align}
  Thus, $g(x)$ is a pdf.
  \item A truncated discrete distribution is one in which a particular class
  cannot be observed and is eliminated from the sample space. In particular, if
  $X$ has range $0,1,2,\ldots$ and the $0$ class cannot be observed (as is
  usually the case), the $0$-truncated random variable $X_T$ has pmf:
  $$P(X_T=x)=\frac{P(X=x)}{P(X>0)}, \quad x=1,2,\ldots$$
  Find the pmf, mean and variance of the $0$-truncated $Poisson(\lambda)$ random
  variable.\\
  The $Poisson(\lambda)$ is defined as
  $P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!}$ for $x=0,1,2,\ldots$. The
  $0$-truncated $Poisson(\lambda)$ is
  $P(X_T=x)=\frac{P(X=x)}{P(X>0)}=\frac{P(X=x)}{1-P(X=0)}$ for $x=1,2,\ldots$,
  because $\sum_{0}^{\infty}P(X=x)=1$.
  $$P(X=0)=\frac{e^{-\lambda}\lambda^0}{0!}=e^{-\lambda}$$
  So:
  $$P(X_T=x)=\frac{e^{-\lambda}\lambda^x}{x!(1-e^{-\lambda})}$$
  The mean of $X$ is:
  $$EX_T=\sum_{x=1}^{\infty}x\frac{e^{-\lambda}\lambda^x}{x!(1-e^{-\lambda})}=\frac{1}{1-e^{-\lambda}}\sum_{x=1}^{\infty}x\frac{e^{-\lambda}\lambda^x}{x!}=\frac{1}{1-e^{-\lambda}}EX=\frac{\lambda}{1-e^{-\lambda}}$$
  Also:
  $$VarX_T=EX_T^2-(EX_T)^2$$
  But:
  $$EX_T^2=\sum_{x=1}{\infty}x^2\frac{e^{-\lambda}\lambda^x}{x!(1-e^{-\lambda})}=\frac{1}{1-e^{-\lambda}}EX^2=\frac{\lambda^2-\lambda}{1-e^{-\lambda}}$$
  Thus:
  $$VarX_T=\frac{\lambda^2-\lambda}{1-e^{-\lambda}}-(\frac{\lambda}{1-e^{-\lambda}})^2$$
\end{enumerate}

\end{document}
